---
title: "COVID-19 patients, deaths, and death rate"
date: "05/22/2020"
output:
  pdf_document: default
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: united
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("~/Documents/Projects/covid19/timeseries/JHU-2019nCoV-to-pandas-DF")
require(ggplot2)
require(reshape2)
require(scales)
require(ggrepel)
```

# Source of the data

The data is compiled from https://github.com/CSSEGISandData/COVID-19.

The python script _script.py_ is a modification of https://github.com/JacopoPan/JHU-2019nCoV-to-pandas-DF


Here I avoided any data modeling and am just showing the raw number of diagnosed patients, deaths, death rate, new cases by day, and new deaths by day.


# Top countries by cases and deaths

```{r fig.height=10, fig.width=15}
deathmat = read.csv("deaths.tsv", header=T,
                    sep="\t", check.names=F, row.names=1)
countdeath = apply(deathmat[nrow(deathmat),1:(ncol(deathmat) - 1)], 2, sum)

casemat = read.csv("confirmed.tsv",
                   header=T, sep="\t", check.names=F, row.names=1)
countcases = apply(casemat[nrow(casemat),1:(ncol(casemat) - 1)], 2, sum)

totaldf = data.frame(Countries=c(names(countdeath), names(countcases)),
                       Values=as.numeric(c(countdeath, countcases)),
                       Variable=c(rep("Deaths", length(countdeath)),
                                  rep("Diagnosed", length(countcases))))
select_countries = names(sort(countcases, TRUE))[1:20]

totaldf = totaldf[totaldf$Countries %in% select_countries, ]
totaldf$Countries = factor(totaldf$Countries, levels=select_countries)

P = ggplot(totaldf, aes(x=Countries, y=Values, fill=Variable)) +
  geom_bar(stat="identity", position="dodge") +
  facet_wrap(~Variable, nrow=2, scales="free") +
  theme_bw(base_size=18) +
  scale_y_log10("Number of patients", labels=comma) +
  theme(legend.position="bottom",
        axis.text.x=element_text(angle=45, hjust=1)) +
  annotation_logticks(sides="lr")
plot(P)
```


## Calculate doubling rate for each country since May 1st

* Red line will show daily increase of twice.

* Purple line will show daily increase of 10%

* Blue line will show daily increase of 1%



```{r}
get_expected = function(datadf, rate=2, init=100){
  list.data = lapply(unique(datadf$variable), function(country){
    tempdf = datadf[datadf$variable==country, ]
    # idx_st = min(which(tempdf$value >= init))
    idx_st = min(which(tempdf$date >= as.Date("2020-05-01")))
    init_val = tempdf$value[idx_st]
    advals = sapply(1:nrow(tempdf), function(x){
      if(x > idx_st){
        return(init_val * (rate**(x - idx_st)))
      }else{
        return(0)
      }
    })
    ad_df = data.frame(variable=country, Rate=rate,
                       date=tempdf$date, value=advals)
    return(ad_df)
  })
  out_df = do.call("rbind", list.data)
  return(out_df)
}
```


# Total morbidity


```{r fig.height = 7, fig.width = 13}
deathmat = read.csv("deaths.tsv", header=T,
                    sep="\t", check.names=F, row.names=1)
countries = c("China", "Iran", "Italy",
              "US", "Russia", "Spain",
              "France", "Canada", "Germany",
              "Korea, South", "United Kingdom", "Israel")
deathmat = deathmat[,c(countries, "date")]

moltendf = melt(deathmat)
moltendf$date = as.Date(moltendf$date)

doubling_df = get_expected(moltendf, rate=2, init=100)
onepercent_df = get_expected(moltendf, rate=1.01, init=100)
tenpercent_df = get_expected(moltendf, rate=1.1, init=100)

P2 = ggplot(moltendf[moltendf$value > 0,],
            aes(x=date, y=(value),
                group=variable)) +
  geom_line() +
  geom_line(data=doubling_df[doubling_df$value > 0,], colour="red3", linetype=2) +
  geom_line(data=onepercent_df[onepercent_df$value > 0,], colour="blue", linetype=3) +
  geom_line(data=tenpercent_df[tenpercent_df$value > 0,], colour="purple", linetype=4) +
  theme_bw(base_size=14) +
  facet_wrap(~variable) +
  scale_y_log10(
    name="Total deaths caused by SARS-CoV-2", labels=comma, limits=c(1, max(moltendf$value))) +
  xlab("Date as of outbreak") +
  ggtitle(paste("Last update on", Sys.time())) +
  annotation_logticks()

plot(P2)
```


# New deaths by day 

```{r fig.height = 7, fig.width = 13}
list.data = lapply(colnames(deathmat)[1:(ncol(deathmat)-1)], function(country){
  new_cases = diff(deathmat[,country])
  addf = data.frame(date=deathmat$date[2:nrow(deathmat)],
                    variable=country,
                    value=new_cases)
})
moltendf = do.call("rbind", list.data)

moltendf$date = as.Date(moltendf$date)

P = ggplot(moltendf, aes(x=date, y=value, group=variable)) +
  geom_line() +
  theme_bw(base_size=14) +
  facet_wrap(~variable, scales="free") +
  scale_y_continuous(
    name="New deaths caused by SARS-CoV-2", labels=comma) +
  ggtitle(paste("Last update on", Sys.time())) +
  xlab("Date as of outbreak")
plot(P)

```

# Total COVID-19 patients

```{r fig.height = 7, fig.width = 13}
casemat = read.csv("confirmed.tsv",
                   header=T, sep="\t", check.names=F, row.names=1)

casemat = casemat[,c(countries, "date")]

moltendf = melt(casemat)
moltendf$date = as.Date(moltendf$date)

doubling_df = get_expected(moltendf, rate=2, init=100)
onepercent_df = get_expected(moltendf, rate=1.01, init=100)
tenpercent_df = get_expected(moltendf, rate=1.1, init=100)


P2 = ggplot(moltendf[moltendf$value > 0, ], aes(x=date, y=(value), group=variable)) +
  geom_line() +
  theme_bw(base_size=14) +
  facet_wrap(~variable) +
  geom_line(data=doubling_df[doubling_df$value > 0,], colour="red3", linetype=2) +
  geom_line(data=onepercent_df[onepercent_df$value > 0,], colour="blue", linetype=3) +
  geom_line(data=tenpercent_df[tenpercent_df$value > 0,], colour="purple", linetype=4) +
  scale_y_log10(
    name="Total COVID-19 patients", labels=comma, limits=c(1, max(moltendf$value))) +
  xlab("Date as of outbreak") +
  ggtitle(paste("Last update on", Sys.time())) +
  annotation_logticks()  

plot(P2)
```


## New COVID-19 patients by day 

```{r fig.height = 7, fig.width = 13}
list.data = lapply(colnames(casemat)[1:(ncol(casemat)-1)], function(country){
  new_cases = diff(casemat[,country])
  addf = data.frame(date=casemat$date[2:nrow(casemat)],
                    variable=country,
                    value=new_cases)
})
moltendf = do.call("rbind", list.data)

moltendf$date = as.Date(moltendf$date)

P = ggplot(moltendf, aes(x=date, y=value, group=variable)) +
  geom_line() +
  theme_bw(base_size=14) +
  facet_wrap(~variable, scales="free") +
  scale_y_continuous(
    name="New COVID-19 patients", labels=comma) +
  ggtitle(paste("Last update on", Sys.time())) +
  xlab("Date as of outbreak")
plot(P)



```


## PCA of countries by total patients

Principal component analysis identifies non-correlating variables (components) from the data and here can identify outlier countries according to a single variable.

In the following plots, I show the first three principal components for total diagnosed patients by day, total deaths by day, and also death rate by day.

```{r fig.height = 5, fig.width = 6}
casemat = read.csv("confirmed.tsv", header=T, sep="\t", check.names=F, row.names=1)

pcamat = casemat[,1:(ncol(casemat) - 1)]
PCA = prcomp(t(pcamat))
pcadf = as.data.frame(PCA$x)
pcadf$Country = colnames(casemat)[1:nrow(pcadf)]
pcadf$Show = ifelse(pcadf$Country %in% countries, TRUE, FALSE)

P = ggplot(pcadf, aes(x=PC1, y=PC2)) +
  geom_point(aes(colour=Show, shape=Show), alpha=0.5) +
  geom_text_repel(data=pcadf[pcadf$Show,], aes(label=Country), colour="blue", size=4) +
  theme_bw(base_size=14) +
  xlab("First principal component") +
  ylab("Second principal component") +
  ggtitle(paste("Last update on", Sys.time())) +
  theme(legend.position="bottom")
plot(P)

P = ggplot(pcadf, aes(x=PC2, y=PC3)) +
  geom_point(aes(colour=Show, shape=Show), alpha=0.5) +
  geom_text_repel(data=pcadf[pcadf$Show,], aes(label=Country), colour="blue", size=4) +
  theme_bw(base_size=14) +
  ylab("Third principal component") +
  xlab("Second principal component") +
  ggtitle(paste("Last update on", Sys.time())) +
  theme(legend.position="bottom")
plot(P)

```

## PCA of countries by morbidity

```{r fig.height = 5, fig.width = 6}
deathmat = read.csv("deaths.tsv", header=T,
                    sep="\t", check.names=F, row.names=1)

pcamat = deathmat[,1:(ncol(deathmat) - 1)]
PCA = prcomp(t(pcamat))
pcadf = as.data.frame(PCA$x)
pcadf$Country = colnames(deathmat)[1:nrow(pcadf)]
pcadf$Show = ifelse(pcadf$Country %in% countries, TRUE, FALSE)

P = ggplot(pcadf, aes(x=PC1, y=PC2)) +
  geom_point(aes(colour=Show, shape=Show), alpha=0.5) +
  geom_text_repel(data=pcadf[pcadf$Show,], aes(label=Country), colour="blue", size=4) +
  theme_bw(base_size=14) +
  xlab("First principal component") +
  ylab("Second principal component") +
  ggtitle(paste("Last update on", Sys.time())) +
  theme(legend.position="bottom")
plot(P)

P = ggplot(pcadf, aes(x=PC2, y=PC3)) +
  geom_point(aes(colour=Show, shape=Show), alpha=0.5) +
  geom_text_repel(data=pcadf[pcadf$Show,], aes(label=Country), colour="blue", size=4) +
  theme_bw(base_size=14) +
  ylab("Third principal component") +
  xlab("Second principal component") +
  ggtitle(paste("Last update on", Sys.time())) +
  theme(legend.position="bottom")
plot(P)

```

## Death rate

The two countries which succeeded in stopping the spread, China and South Korea, provide the most accurate measurement of true death rate.
In the death rate plot below, the range of the death rate of these two countries is shown by pink.

```{r fig.height = 7, fig.width = 13}
deathmat = read.csv("deaths.tsv", header=T,
                    sep="\t", check.names=F, row.names=1)
casemat = read.csv("confirmed.tsv", header=T, sep="\t", check.names=F, row.names=1)
deathrate = deathmat[,1:(ncol(deathmat) - 1)] / (0.1  + casemat[,setdiff(colnames(deathmat), "date")])

deathratedf = deathrate
deathratedf$date = casemat$date
moltendf = melt(deathratedf[,c(countries, "date")])

moltendf$date = as.Date(moltendf$date)

moltendf$Label = NA
deathratedf$date = as.Date(deathratedf$date)

textmat = casemat[which.max(casemat[,"China"]), ]
textdf = melt(as.data.frame(textmat[,c(countries, "date")]))
textdf$date = min(as.Date(deathratedf$date)) + 40
textdf$Label = paste("N =", comma(textdf$value))
textdf$value = 0.08

P = ggplot(moltendf, aes(x=date, y=value, group=variable)) +
  geom_rect(xmin=min(deathratedf$date),
            xmax=max(deathratedf$date+1),
            ymin=max(deathratedf$China),
            ymax=max(deathratedf[,"Korea, South"]),
            alpha=0.05,
            fill="pink") +
  geom_line() +
  geom_text(data=textdf, aes(x=date, y=value, label=Label), colour='blue') +
  theme_bw(base_size=14) +
  facet_wrap(~variable, scales="free") +
  scale_y_continuous(
    name=expression("COVID-19 daily "*frac("deaths", "cases")),
    labels=comma) +
  xlab("Date as of outbreak") +
  ggtitle(paste("Last update on", Sys.time())) +
  coord_cartesian(ylim=c(0, 0.2))
plot(P)
```

```{r fig.height = 5, fig.width = 6}
pcamat = deathrate
PCA = prcomp(t(pcamat))
pcadf = as.data.frame(PCA$x)
pcadf$Country = colnames(deathmat)[1:nrow(pcadf)]
pcadf$Show = ifelse(pcadf$Country %in% countries, TRUE, FALSE)

P = ggplot(pcadf, aes(x=PC1, y=PC2)) +
  geom_point(aes(colour=Show, shape=Show), alpha=0.5) +
  geom_text_repel(data=pcadf[pcadf$Show,], aes(label=Country), colour="blue", size=4) +
  theme_bw(base_size=14) +
  xlab("First principal component") +
  ylab("Second principal component") +
  theme(legend.position="bottom")
plot(P)

P = ggplot(pcadf, aes(x=PC2, y=PC3)) +
  geom_point(aes(colour=Show, shape=Show), alpha=0.5) +
  geom_text_repel(data=pcadf[pcadf$Show,], aes(label=Country), colour="blue", size=4) +
  theme_bw(base_size=14) +
  ylab("Third principal component") +
  xlab("Second principal component") +
  theme(legend.position="bottom")
plot(P)

```


# Canada

```{r fig.height = 13, fig.width = 15}
casepath = "COVID-19/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv"
casedf = read.csv(casepath, header=T, check.names=F)
casedf = casedf[casedf[,2]=="Canada", c(1, 5:ncol(casedf))]
provinces = casedf[apply(casedf[,2:ncol(casedf)], 1, sum) > 0, 1]
casedf = casedf[casedf[,1] %in% provinces,]


moltendf = melt(casedf)
moltendf$date = as.Date(as.character(moltendf$variable), tryFormats="%m/%d/%y")

moltendf$variable = moltendf[,"Province/State"]
doubling_df = get_expected(moltendf, rate=2, init=100)
doubling_df[,"Province/State"] = doubling_df$variable
onepercent_df = get_expected(moltendf, rate=1.01, init=100)
onepercent_df[,"Province/State"] = onepercent_df$variable
tenpercent_df = get_expected(moltendf, rate=1.1, init=100)
tenpercent_df[,"Province/State"] = tenpercent_df$variable


textdf = melt(casedf[,c(1, ncol(casedf))])
textdf$date = as.Date(as.character(textdf$variable), tryFormats="%m/%d/%y")
textdf$date = min(as.Date(textdf$date)) - 50
textdf$Label = paste("N =", comma(round(textdf$value)))
textdf$value = 2000

P = ggplot(moltendf[moltendf$value > 0,], aes(x=date, y=value, group=`Province/State`)) +
  geom_line() +
  geom_line(data=doubling_df[doubling_df$value > 0,], colour="red3", linetype=2) +
  geom_line(data=onepercent_df[onepercent_df$value > 0,], colour="blue", linetype=3) +
  geom_line(data=tenpercent_df[tenpercent_df$value > 0,], colour="purple", linetype=4) +
  geom_text(data=textdf, aes(label=Label), colour="blue") +
  theme_bw(base_size=18) +
  facet_wrap(~`Province/State`) +
  scale_y_log10("Number of patients diagnosed with COVID-19", labels=comma, limits=c(1, max(moltendf$value))) +
  xlab("Date as of global incidence")
plot(P)
```


## Canada's deaths

```{r fig.height = 13, fig.width = 15}
casepath = "COVID-19/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv"
casedf = read.csv(casepath, header=T, check.names=F)
casedf = casedf[casedf[,2]=="Canada", c(1, 5:ncol(casedf))]
provinces = casedf[apply(casedf[,2:ncol(casedf)], 1, sum) > 0, 1]
casedf = casedf[casedf[,1] %in% provinces,]


moltendf = melt(casedf)
moltendf$date = as.Date(as.character(moltendf$variable), tryFormats="%m/%d/%y")

moltendf$variable = moltendf[,"Province/State"]
doubling_df = get_expected(moltendf, rate=2, init=100)
doubling_df[,"Province/State"] = doubling_df$variable
onepercent_df = get_expected(moltendf, rate=1.01, init=100)
onepercent_df[,"Province/State"] = onepercent_df$variable
tenpercent_df = get_expected(moltendf, rate=1.1, init=100)
tenpercent_df[,"Province/State"] = tenpercent_df$variable


textdf = melt(casedf[,c(1, ncol(casedf))])
textdf$date = as.Date(as.character(textdf$variable), tryFormats="%m/%d/%y")
textdf$date = min(as.Date(textdf$date)) - 5
textdf$Label = paste("N =", comma(round(textdf$value)))
textdf$value = max(moltendf$value)

P = ggplot(moltendf[moltendf$value > 0,], aes(x=date, y=value, group=`Province/State`)) +
  geom_line() +
  geom_line(data=doubling_df[doubling_df$value > 0,], colour="red3", linetype=2) +
  geom_line(data=onepercent_df[onepercent_df$value > 0,], colour="blue", linetype=3) +
  geom_line(data=tenpercent_df[tenpercent_df$value > 0,], colour="purple", linetype=4) +
  geom_text(data=textdf, aes(label=Label), colour="blue") +
  theme_bw(base_size=18) +
  facet_wrap(~`Province/State`) +
  scale_y_log10("Number of deaths due to COVID-19", labels=comma, limits=c(1, max(moltendf$value))) +
  xlab("Date as of first Canadian death")
plot(P)
```
